Genome-wide changes in expression during inflorescence architecture development

### 2015-10-09 ms.Rmd
Fix table numbering Hormones, other writing

---
title: Genome-wide changes in expression during inflorescence architecture development
  in rice
csl: csl/the-plant-journal.csl
header-includes:
  \usepackage{graphicx}
  \usepackage[utf8]{inputenc}
  \DeclareUnicodeCharacter{2265}{$\geq$}
output:
  word_document:
    fig_caption: yes
    reference_docx: csl/wordStyles.docx
  BiocStyle::html_document:
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
    dev: "CairoPDF"
    toc: true
  pdf_document:
    fig_caption: yes
    number_sections: yes
  rmarkdown::tufte_handout:
    default
bibliography: bib/papersLib.bibtex
---

@@

The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes suggests that the clusters capture biologically meaningful groups of genes. Two G1-Like (G1L/ALOG) genes, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively. Only three ALOG genes are expressed in the meristem RNA sequencing dataset, but they share a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`). Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff]. *G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff]. This might be explained by redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5* [**This is just speculation!** I checked and there is some homology between the RNAi construct and other ALOG genes, but I'm not sure how much is required for off-target effects.]

The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in flower development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions. Two other AP2 genes, *INDETERMINATE SPIKELET*, which acts redundantly with *SNB* to promote transition from SM to FM [@Lee:2012hj], and *FRIZZY PANICLE*, which is required for determination of SM [**Komatsu 2003 FZP paper**], were found in clusters containing genes that are highly expressed in the SM.

@@

The strong expression of SPL genes (which possess the SBP-box) in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development. Overexpression of *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14*), which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. *WFP*, *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed (**ref: http://www.ncbi.nlm.nih.gov/pmc/articles/PMC1557610/**).

These patterns are not intended to represent each member of each gene family: for example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM. An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM. *PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs, and mutants of the former gene produce fewer spikelets than wild-type plants [@Gao:2010iz] (**check role of PAP2**). In contrast to *PAP2*, *LHS1* was detected only in the SM samples, indicative of the overall pattern for other MADS genes.

@@

The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants. In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen (recently reviewed by **ref: Prunet/Jack review**). In rice, functions homologous to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on? check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**]. [**Add info about tomato MADS**].

@@

The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db; @Durinck:2009ki]. An initial alignment of all proteins was performed with `Clustal Omega` using a full distance matrix (**ref clustalo**). To clean the alignment, poorly conserved regions where greater than `r minpcnongap` percent of sequences had gaps or less than `r minpcident` percent of pairwise comparisons between sequences had identical residues were removed. To recover MADS genes related to the Arabidopsis MIKC clade, a UGPMA tree was estimated from the cleaned alignment and the tree was cut at the lowest height that returned a clade including all type II Arabidopsis MADS proteins listed in the Arabidopsis transcription factor database on the Arabidopsis Gene Regulatory Information Server (**ref for AGRIS**), with as few type I proteins as possible. Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package (**ref:ape***) rooted to type I MADS sequences (`r og`). L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same method described above for the rice samples. The tree was plotted using the `ggtree` extension for `ggplot2` (**ref ggtree**).

@@

**`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**. Expression of rice G1/ALOG family genes in inflorescence meristems. Genes were retrieved by querying Phytozome 10.3 for the Pfam domain PF04852 [@Goodstein:2012ho] (**ref: Pfam**). Red or blue points indicate whether the gene was detected or not detected respectively for each library. Only three G1 genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA*) were detected, but they share a similar pattern of expression.

>>>>

---
title: Genome-wide changes in expression during inflorescence architecture development
  in rice
csl: csl/the-plant-journal.csl
header-includes: \usepackage{graphicx} \usepackage[utf8]{inputenc} \DeclareUnicodeCharacter{2265}{$\geq$}
output:
  BiocStyle::html_document:
    dev: CairoPDF
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
    toc: yes
  pdf_document:
    fig_caption: yes
    number_sections: yes
  rmarkdown::tufte_handout: default
  word_document:
    fig_caption: yes
    reference_docx: csl/wordStyles.docx
bibliography: bib/papersLib.bibtex
---

@@

Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). The latter,

Two homologs of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*) and *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*), are present in cluster `r getClusterName('inc', 'inc')`. The LOG enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log* mutants produce small panicles with small SAMs and branching defects, and have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes [@Kurakawa:2007go]. In contrast, the LABA1 protein present in *O. rufipogon*, which has an identical primary sequence to the protein from *O. sativa* ssp. *japonica* cv. Nipponbare, also activates cytokinin *in vitro* [**ref Gu 2015; Hua et all 2015**], and introduction of the *Laba1* allele from *O. rufipogon* into an *O. sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [**Hua et all 2015**], but reduces the number of grains per panicle and the number of tillers per plant [**ref Gu 2015**], suggesting a link between increased cytokinin signalling and reduced panicle branching. *laba1* is a domestication allele found at high frequencies in cultivated accessions, suggesting that it was affected by artificial selection either for awn phenotype or grain yield [**ref Gu 2015; Hua et all 2015**]. No reports of *LOGL1* function in rice were found. An uncharecterized gene annotated as a cytokinin dehydrogenase, *LOC_Os05g31040*, is present in the complementary cluster (cluster `r getClusterName('dec', 'dec')`), which suggests a model of increasing cytokinin activation and decreasing inactivation from RM–SM, but it is currently not clear how such a mechanism might influence meristem identity. There are also genes that may be involved in cytokinin signalling in other clusters, including a type-A response regulator (*RR3*) and a cytokinin receptor (*PHP2*) in cluster `r getClusterName('dec', 'small')`, containing genes more highly expressed in RM, suggesting that the role of cytokinins in inflorescence development may be complex. **Grain number 1a?**. **Effect of cytokinin on meristem size/activity?**

The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes suggests that the clusters capture biologically meaningful groups of genes. Two G1-Like (G1L/ALOG) genes, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively. Only three G1L genes are expressed in the meristem RNA sequencing dataset, but they share a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`). Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff]. *G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff]. This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5* [**This is just speculation!** I checked and there is some homology between the RNAi construct and other G1L genes, but I'm not sure how much is required for off-target effects.]

The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in flower development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions. Two other AP2 genes, *INDETERMINATE SPIKELET*, which acts redundantly with *SNB* to promote transition from SM to FM [@Lee:2012hj], and *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], were found in clusters containing genes that are highly expressed in the SM.

@@

The strong expression of SPL genes (which possess the SBP-box) in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development. Overexpression of *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14*), which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. *WFP*, *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc].

These patterns are not intended to represent each member of each gene family. For example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM. An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM. *PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *PAP2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz, @Kobayashi:2010hn]. In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [], was detected only in the SM samples and is thus representative the overall enrichment of other MADS genes in those samples [**ref Jeon 2000, Agrawal 2005**].

@@

The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants. In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen (recently reviewed by @Prunet:2013ki). In rice, functions homologous to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on? check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**]. [**Add info about tomato MADS**].

@@

The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db; @Durinck:2009ki]. An initial alignment of all proteins was performed with `Clustal Omega` using a full distance matrix [@Sievers:2011fn]. To clean the alignment, poorly conserved regions where greater than `r minpcnongap` percent of sequences had gaps or less than `r minpcident` percent of pairwise comparisons between sequences had identical residues were removed. To recover MADS genes related to the Arabidopsis MIKC clade, a UGPMA tree was estimated from the cleaned alignment and the tree was cut at the lowest height that returned a clade including all type II Arabidopsis MADS proteins listed in the Arabidopsis transcription factor database on the Arabidopsis Gene Regulatory Information Server [@Yilmaz:2011ew], with as few type I proteins as possible. Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package [@Paradis:2004dv] rooted to type I MADS sequences (`r og`). L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same method described above for the rice samples. The tree was plotted using the `ggtree` extension for `ggplot2` (**ref ggtree**).

@@

**`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**. Expression of rice G1/ALOG family genes in inflorescence meristems. Genes were retrieved by querying Phytozome 10.3 for the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj]. Red or blue points indicate whether the gene was detected or not detected respectively for each library. Only three G1 genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA*) were detected, but they share a similar pattern of expression.
